Distinct smell and taste disorder phenotype of post-acute COVID-19 sequelae

### 2022-07-13 paper/markdown/supplementary_material.Rmd
Figure content change
Figure content changed (less supplements...), changes in the legend and supplement text. Update of the figure-making scripts to the newest version of figur package

---
title: "Distinct smell and taste disorder phenotype of post-acute COVID-19 sequelae"
subtitle: "Supplementary Material"
author: "HACT and CovILD study teams"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

The complete list of study variables and stratification scheme is provided in __Supplementary Table S\@ref(tab:tab-hact-vars)__ for the HACT study and in __Supplementary Table S\@ref(tab:tab-covild-vars)__ for the CovILD study.

@@

A total of `r length(globals$hact_symptoms)` was surveyed in the HACT study cohorts (__Supplementary Figure S\@ref(fig:fig-hact-freq)__ and __Supplementary Table S\@ref(tab:tab-hact-vars)__). The symptom duration was coded as follows: absent: 0 days, 1 - 3 days: 3 days, up to 1 week: 7 days, up to 2 weeks: 14 days, up to 4 weeks: 28 days, up to 3 months: 90 days, up to 6 months: 90 days, over 6 months: 90 days. Acute symptoms were defined as complaints present during the first 14 days after clinical onset of COVID-19. Long COVID symptoms were defined as complaints present for $\geq$ 28 days. PASC symptoms were defined as manifestations present for $\geq$ 90 days [@Venkatesan2021; @Shah2021].

@@

### Rating of physical recovery, mental health and quality of life in the HACT study

@@

where $(1|individual)$ indicates the random effect of the individual and $time$ and $time^2$ indicate the first- and second-order time effect terms. The first-order term estimate was interpreted as a measure of the recovery speed and the second-order term estimate was used to assess the plateau/rebound effect. Significance of the accuracy gain of the full second-order model compared with the nested null model was determined by likelihood ratio test (LRT) versus the nested first-order and null models, respectively. Likelihood ratio $\lambda$ statistic (full versus null model) was used as an effect size measure. 
Individuals from the HACT or CovILD studies with the complete longitudinal symptom record were included in the kinetic modeling tasks. Results of the kinetic modeling were adjusted for multiple comparisons with Benjamini-Hochberg method [@Benjamini1995].

@@

To assess co-occurrence or exclusivity of symptoms, simple matching distances between manifestations of each acute COVID-19, long COVID and PASC in the HACT study cohorts were calculated (package _scrime_ and development package _clustTools_ [https://github.com/PiotrTymoszuk/clustTools]) [@Boriah2008; @Schwender2018]. Subsequently, the distance matrix was subjected to multi-dimensional scaling (MDS, k = 2 dimensions, package _stats_, function _cmdscale()_). Association of specific symptoms was assessed by visual analysis of MDS coordinate plots.

### Apriori analysis of acute COVID-19, long COVID and PASC symptoms in the HACT study

Frequent combinations of symptoms of long COVID and PASC in the HACT study cohorts were identified with the apriori algorithm (package _arules_) [@Agrawal1993; @Hahsler2005] with the minimal support cutoff of 0.1, 2 - 10 item transaction length, confidence > 0.8 and lift > 2. The support statistic were used to estimate the symptom combination frequency. The confidence value was treated as an estimate of conditional probability of the symptom co-occurrence. The lift statistic was interpreted as a measure of the symptom dependence (lift = 1, symptoms are independent). The complete analysis results are presented in __Supplementary Tables S\@ref(tab:tab-apriori-long)__ and __S\@ref(tab:tab-apriori-pasc)__.

@@

COVID-19 recovery clusters of the training Autria (AT) HACT cohort participants in respect to symptom-specific recovery times (__Figure 1A__) were defined with the PAM (partitioning around medoids) algorithm and Euclidean distance statistic (packages _cluster_, _philentropy_ and development package _clustTools_ [https://github.com/PiotrTymoszuk/clustTools]) [@Drost2018; @Schubert2019]. The set of participants with the complete clustering variable set was included in the analysis. The symptom recovery times were not subjected to any type of pre-processing. The choice of the clustering procedure was motivated by the analysis of the clustering variance (ratio of the total between-cluster to total sum of squares) and clustering structure stability in 10-fold cross-validation (metric: rate of correct cluster assignment, cluster assignment predicted by k = 5 nearest neighbors label propagation algorithm, package _clustTools_) [@Lange2004; @Leng2014] for several clustering algorithms as presented in __Supplementary Figure S\@ref(fig:fig-clust-devel)A__. The optimal number of clusters was determined by the bend of the total within-cluster sum of squares curve (__Supplementary Figure S\@ref(fig:fig-clust-devel)B__, package _factoextra_) [@Kassambara2020]. 
Permutation importance of specific clustering variables was investigated by calculating difference in clustering variance (ratio of total between-cluster sum of squares to total sum of squares) between the initial clustering object and the clustering object with the given variable reshuffled at random (package _clustTools_). 
Assignment of the Italy HACT cohort participants to the recovery clusters was accomplished with k-nearest neighbors label propagation algorithm (k = 5) [@Leng2014]. The clustering efficacy in the training AT cohort and the test IT cohort measured by clustering variance statistic defined above was similar (AT: `r clustTools::var(part_clust$clust_north)$frac_var %>% signif(2)`, IT: `r clustTools::var(part_clust$clust_south)$frac_var %>% signif(2)`).

@@

```{r tab-hact-vars, tab.cap = 'HACT study variables.'}

suppl_tbl$hact_vars %>% 
  select(- Stratification) %>% 
  flextable %>% 
  width(1, width = 3.7, unit = 'cm') %>% 
  width(2, width = 3.9, unit = 'cm') %>% 
  width(3, width = 1.8, unit = 'cm')  %>% 
  width(4, width = 8.7, unit = 'cm') %>% 
  theme_vanilla %>% 
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

@@

suppl_tbl$covild_vars %>% 
  flextable %>% 
  width(1, width = 4.5, unit = 'cm') %>% 
  width(2, width = 4.5, unit = 'cm') %>% 
  width(3, width = 2, unit = 'cm')  %>% 
  theme_vanilla %>% 
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

@@

```{r tab-apriori-long, tab.cap = 'Results of apriori analysis for long COVID-19 symptoms in the HACT study . Top 10 most frequent symptom combinations (highest support value) in the AT (Austria) and IT (Italy) cohort are presented. The full results are available as a supplementary Excel table.'}

suppl_tbl$apriori_long %>% 
  group_by(Cohort) %>% 
  top_n(10, Support) %>% 
  ungroup %>% 
  arrange(Cohort, Support) %>% 
  select(-Time) %>% 
  flextable %>% 
  width(1, width = 1.5, unit = 'cm') %>% 
  width(2, width = 9.5, unit = 'cm') %>% 
  width(3:5, width = 2.2, unit = 'cm')  %>% 
  theme_vanilla %>% 
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

@@

```{r tab-apriori-pasc, tab.cap = 'Results of apriori analysis for PASC symptoms in the HACT study. Top 10 most frequent symptom combinations (highest support value) in the AT (Austria) and IT (Italy) cohort are presented. The full results are available as a supplementary Excel table.'}

suppl_tbl$apriori_pasc %>% 
  group_by(Cohort) %>% 
  top_n(10, Support) %>% 
  ungroup %>% 
  arrange(Cohort, Support) %>% 
  select(-Time) %>% 
  flextable %>% 
  width(1, width = 1.5, unit = 'cm') %>% 
  width(2, width = 9.5, unit = 'cm') %>% 
  width(3:5, width = 2.2, unit = 'cm')  %>% 
  theme_vanilla %>% 
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

@@

```{r tab-clust-base-at, tab.cap = 'Demographic and baseline clinical characteristic at the COVID-19 onset of the HACT study participants assigned to the recovery clusters, Austria (AT) cohort.'}

suppl_tbl$baseline_clusters_north %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('BMI: body mass index, overweight > 25 kg/m\u00B2, obesity > 30 kg/m\u00B2,'), ref_symbols = 'b', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('Frequent respiratory infections, > 2 per year.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(16, 1, value = as_paragraph('Frequent bacterial infections with antibiotic therapy, > 2 per year.'), ref_symbols = 'd', part = 'body') %>% 
  theme_vanilla %>%
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

@@

```{r tab-clust-base-it, tab.cap = 'Demographic and baseline clinical characteristic at the COVID-19 onset of the HACT study participants assigned to the recovery clusters, Italy (IT) cohort.'}

suppl_tbl$baseline_clusters_south %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('BMI: body mass index, overweight > 25 kg/m\u00B2, obesity > 30 kg/m\u00B2,'), ref_symbols = 'b', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('Frequent respiratory infections, > 2 per year.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(16, 1, value = as_paragraph('Frequent bacterial infections with antibiotic therapy, > 2 per year.'), ref_symbols = 'd', part = 'body') %>% 
  theme_vanilla %>%
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

```{r tab-clust-reco-at, tab.cap = 'COVID-19 course and recovery in the HACT study participants assigned to the recovery clusters, Austria (AT) cohort.'}

suppl_tbl$cov_clusters_north %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('Number of acute COVID-19 symptoms.'), ref_symbols = 'b', part = 'body') %>% 
  footnote(4, 1, value = as_paragraph('Number of long COVID symptoms.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(7, 1, value = as_paragraph('Self-reported incomplete recovery.'), ref_symbols = 'd', part = 'body') %>% 
  footnote(11, 1, value = as_paragraph('ANX: anxiety.'), ref_symbols = 'e', part = 'body') %>% 
  footnote(12, 1, value = as_paragraph('DPR: depression.'), ref_symbols = 'f', part = 'body') %>% 
  footnote(14, 1, value = as_paragraph('OMH score: overall mental health impairment score'), ref_symbols = 'g', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('QoL score: quality of life impairment score'), ref_symbols = 'h', part = 'body') %>%
  theme_vanilla %>%
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

```{r tab-clust-reco-it, tab.cap = 'COVID-19 course and recovery in the HACT study participants assigned to the recovery clusters, Italy (AT) cohort.'}

suppl_tbl$cov_clusters_south %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('Number of acute COVID-19 symptoms.'), ref_symbols = 'b', part = 'body') %>% 
  footnote(4, 1, value = as_paragraph('Number of long COVID symptoms.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(7, 1, value = as_paragraph('Self-reported incomplete recovery.'), ref_symbols = 'd', part = 'body') %>% 
  footnote(11, 1, value = as_paragraph('ANX: anxiety.'), ref_symbols = 'e', part = 'body') %>% 
  footnote(12, 1, value = as_paragraph('DPR: depression.'), ref_symbols = 'f', part = 'body') %>% 
  footnote(14, 1, value = as_paragraph('OMH score: overall mental health impairment score'), ref_symbols = 'g', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('QoL score: quality of life impairment score'), ref_symbols = 'h', part = 'body') %>%
  theme_vanilla %>%
  font(part = 'all', fontname = 'Cambria') %>% 
  fontsize(size = 10, part = 'all')

@@

```{r fig-hact-freq, fig.width = mm_inch(suppl$hact_freq$w), fig.height = mm_inch(suppl$hact_freq$h), fig.cap = 'Frequency of COVID-19 symptoms in acute COVID-19, long COVID and PASC in the HACT study.'}

@@

__Supplementary Figure S\@ref(fig:fig-hact-freq). Frequency of COVID-19 symptoms in acute COVID-19, long COVID and PASC in the HACT study.__

Frequency of symptoms of acute COVID-19 (first 14 days post clinical onset), long COVID ($\geq$ 28 days) and PASC (post-acute sequelae of COVID-19, $\geq$ 90 days) in the AT (Austria, __A__) and IT (Italy, __B__) HACT study cohorts expressed as percentages of the cohort. Point size and color represents the percentage. Numbers of complete observations are indicated under the plot.

Dim. appetite: diminished appetite, Imp. concentration: impaired concentration, Imp. walk: impaired walk, Imp. FMS: impaired fine motor skills.

@@

```{r fig-hact-recovery, fig.width = mm_inch(suppl$hact_recovery$w), fig.height = mm_inch(suppl$hact_recovery$h), fig.cap = 'Kinetic of recovery from leading acute COVID-19 symptoms in the HACT study.'}

@@

__Supplementary Figure S\@ref(fig:fig-hact-recovery). Kinetic of recovery from leading acute COVID-19 symptoms in the HACT study.__

Percentages of individuals with fever (__A__), diminished appetite (__B__), joint pain (__C__), muscle pain (__D__), fatigue (__E__) and tachypnea (__F__) in the AT (Austria) and IT (Italy) HACT study cohorts at particular time points after clinical onset. Numbers of complete observations are indicated under the plots.

@@

Frequency of symptoms during acute COVID-19 and at the 60-, 100-, 180- and 360-day follow-ups in ambulatory (__A__), moderate (__B__) and severe COVID-19 (__C__) participants of the CovILD study expressed as percentages of individuals with the complete longitudinal data set. Point size and color represents the percentage. Numbers of complete observations are indicated under the plots.

@@

Percentages of individuals with the complete longitudinal data set suffering from smell disorders (__A__), reduced physical performance (__B__) and dyspnea (__C__) in the ambulatory, moderate and severe COVID-19 subsets of the CovILD study during acute COVID-19 and at the 60-, 100-, 180- and 360-day follow-ups. Numbers of complete observations are indicated under the plots.

@@

Association of self-reported and sniffing stick test hyposmia rates was investigated with Cohen's $\kappa$ statistic. Statistical significance was assessed with Wald's Z test corrected for multiple testing with Benjamini-Hochberg method. Percentages of individuals with self-reported and test hyposmia within the entire cohort (__A__), the ambulatory (__B__), moderate (__C__) and severe (__D__) COVID-19 subsets of the CovILD-19 subsets are presented in bubble plots. Point size and color represents the percentage. $\kappa$ values with 95$\%$ confidence intervals, p values and numbers of complete observations are indicated in the plot captions.

@@

```{r fig-mds-acute, fig.width = mm_inch(suppl$mds_acute$w), fig.height = mm_inch(suppl$mds_acute$h), fig.cap = 'Multi-dimensional scaling analysis of acute COVID-19 symptoms in the HACT study.'}

@@

__Supplementary Figure S\@ref(fig:fig-mds-acute). Multi-dimensional scaling analysis of acute COVID-19 symptoms in the HACT study.__

Binary symptom occurrence data for acute COVID-19 (first 14 days after clinical onset) in the HACT study Austria (AT) and Italy (IT) cohorts were subjected to two-dimensional multi-dimensional scaling (MDS) with simple matching distance (SMD) between the symptoms. MDS coordinates are presented in point plots. Selected data points are labeled with the symptom names. Percentages of the data set variance associated with the MDS dimensions are indicated in the plot axes. Numbers of complete observations are indicated under the plots.

@@

```{r fig-clust-devel, fig.width = mm_inch(suppl$clust_devel$w), fig.height = mm_inch(suppl$clust_devel$h), fig.cap = 'Definition of the COVID-19 recovery clusters and clustering feature importance in the HACT study.'}

@@

__Supplementary Figure S\@ref(fig:fig-clust-devel). Definition of the COVID-19 recovery clusters and clustering feature importance in the HACT study.__

Individuals of the training Austria (AT) cohort of the HACT study were subjected to clustering in respect to symptom-specific recovery times with the PAM (partitioning around medoids) algorithm and Euclidean distance measure.

__(A)__ Comparison of performance of various algorithms (HCl: hierarchical clustering, SOM + HCl: combined self-organizing map and hierarchical clustering, k-means) and distance statistic in clustering of the training data set investigated by clustering variance (ratio of total between-cluster sum of squares to total sum of squares) and cluster assignment accuracy in 10-fold cross-validation (CV).

__(B)__ Determination of the optimal cluster number in the PAM clustering of the training cohort by the bend of the total within-cluster sum of squares curve.

__(C)__ Permutation importance of the clustering features (symptoms) for clustering of the training cohort expressed as the difference in clustering variance (ratio of total between-cluster sum of squares to total sum of squares) between the initial clustering object and the clustering object with the given variable reshuffled at random.

@@

```{r fig-cov-clusters, fig.width = mm_inch(suppl$cov_clusters$w), fig.height = mm_inch(suppl$cov_clusters$h), fig.cap = 'Numbers of acute COVID-19 and long COVID symptoms in the HACT study recovery clusters.'}

>>>>

---
title: "Distinct smell and taste disorder phenotype of post-acute COVID-19 sequelae"
subtitle: "Supplementary Material"
author: "Hralth after COVID-19 in Tyrol and CovILD study teams"
date: "`r format(Sys.time(), '%Y-%m-%d')`"

@@

The complete list of study variables and stratification scheme is provided in __Supplementary Table S\@ref(tab:tab-hact-vars)__ for the Health after COVID-19 in Tyrol survey study and in __Supplementary Table S\@ref(tab:tab-covild-vars)__ for the CovILD cohort.

@@

A total of `r length(globals$hact_symptoms)` was surveyed in the survey study cohorts (__Supplementary Figure S\@ref(fig:fig-hact-freq)__ and __Supplementary Table S\@ref(tab:tab-hact-vars)__). The symptom duration was coded as follows: absent: 0 days, 1 - 3 days: 3 days, up to 1 week: 7 days, up to 2 weeks: 14 days, up to 4 weeks: 28 days, up to 3 months: 30 moths, up to 6 months: 3 months, over 6 months: 3 months. Acute symptoms were defined as complaints present during the first 14 days after clinical onset of COVID-19.

@@

### Rating of physical recovery, mental health and quality of life in the survey study

@@

where $(1|individual)$ indicates the random effect of the individual and $time$ and $time^2$ indicate the first- and second-order time effect terms. The first-order term estimate was interpreted as a measure of the recovery speed and the second-order term estimate was used to assess the plateau/rebound effect. Significance of the accuracy gain of the full second-order model compared with the nested null model was determined by likelihood ratio test (LRT) versus the nested first-order and null models, respectively. Likelihood ratio $\lambda$ statistic (full versus null model) was used as an effect size measure. 
Individuals from the survey study or the CovILD cohort with the complete longitudinal symptom record were included in the kinetic modeling tasks. Results of the kinetic modeling were adjusted for multiple comparisons with Benjamini-Hochberg method [@Benjamini1995].

@@

To assess co-occurrence or exclusivity of symptoms, simple matching distances between manifestations during the first 14 days, at $\geq$ 28 days and at $\geq$ 3 months after clinical onset in the survey study cohorts were calculated (package _scrime_ and development package _clustTools_ [https://github.com/PiotrTymoszuk/clustTools]) [@Boriah2008; @Schwender2018]. Subsequently, the distance matrix was subjected to multi-dimensional scaling (MDS, k = 2 dimensions, package _stats_, function _cmdscale()_). Association of specific symptoms was assessed by visual analysis of MDS coordinate plots.

### Apriori analysis of COVID-19 symptoms in the survey study

Frequent combinations of symptoms during the first 14 days, at $\geq$ 28 days and at $\geq$ 3 months after clinical onset in the survey study cohorts were identified with the apriori algorithm (package _arules_) [@Agrawal1993; @Hahsler2005] with the minimal support cutoff of 0.1, 2 - 10 item transaction length, confidence > 0.8 and lift > 2. The support statistic were used to estimate the symptom combination frequency. The confidence value was treated as an estimate of conditional probability of the symptom co-occurrence. The lift statistic was interpreted as a measure of the symptom dependence (lift = 1, symptoms are independent).

@@

COVID-19 recovery clusters of the training Autria (AT) survey cohort participants in respect to symptom-specific recovery times (__Figure 1A__) were defined with the PAM (partitioning around medoids) algorithm and Euclidean distance statistic (packages _cluster_, _philentropy_ and development package _clustTools_ [https://github.com/PiotrTymoszuk/clustTools]) [@Drost2018; @Schubert2019]. The set of participants with the complete clustering variable set was included in the analysis. The symptom recovery times were not subjected to any type of pre-processing. The choice of the clustering procedure was motivated by the analysis of the clustering variance (ratio of the total between-cluster to total sum of squares) and clustering structure stability in 10-fold cross-validation (metric: rate of correct cluster assignment, cluster assignment predicted by k = 5 nearest neighbors label propagation algorithm, package _clustTools_) [@Lange2004; @Leng2014] for several clustering algorithms as presented in __Supplementary Figure S\@ref(fig:fig-clust-devel)A__. The optimal number of clusters was determined by the bend of the total within-cluster sum of squares curve (__Supplementary Figure S\@ref(fig:fig-clust-devel)B__, package _factoextra_) [@Kassambara2020]. 
Permutation importance of specific clustering variables was investigated by calculating difference in clustering variance (ratio of total between-cluster sum of squares to total sum of squares) between the initial clustering object and the clustering object with the given variable reshuffled at random (package _clustTools_). 
Assignment of the Italy survey cohort participants to the recovery clusters was accomplished with k-nearest neighbors label propagation algorithm (k = 5) [@Leng2014]. The clustering efficacy in the training AT cohort and the test IT cohort measured by clustering variance statistic defined above was similar (AT: `r clustTools::var(part_clust$clust_north)$frac_var %>% signif(2)`, IT: `r clustTools::var(part_clust$clust_south)$frac_var %>% signif(2)`).

@@

```{r tab-hact-vars, tab.cap = 'Survey study variables.'}

suppl_tbl$hact_vars %>% 
  select(- Stratification) %>% 
  flextable %>% 
  width(1, width = 3.7, unit = 'cm') %>% 
  width(2, width = 3.9, unit = 'cm') %>% 
  width(3, width = 1.8, unit = 'cm')  %>% 
  width(4, width = 8.7, unit = 'cm') %>% 
  theme_vanilla

@@

suppl_tbl$covild_vars %>% 
  flextable %>% 
  width(1, width = 4.5, unit = 'cm') %>% 
  width(2, width = 4.5, unit = 'cm') %>% 
  width(3, width = 2, unit = 'cm')  %>% 
  theme_vanilla

@@

```{r tab-clust-base-at, tab.cap = 'Demographic and baseline clinical characteristic at the COVID-19 onset of the survey study participants assigned to the recovery clusters, Austria (AT) cohort.'}

suppl_tbl$baseline_clusters_north %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('BMI: body mass index, overweight > 25 kg/m\u00B2, obesity > 30 kg/m\u00B2,'), ref_symbols = 'b', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('Frequent respiratory infections, > 2 per year.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(16, 1, value = as_paragraph('Frequent bacterial infections with antibiotic therapy, > 2 per year.'), ref_symbols = 'd', part = 'body') %>% 
  theme_vanilla

@@

```{r tab-clust-base-it, tab.cap = 'Demographic and baseline clinical characteristic at the COVID-19 onset of the survey study participants assigned to the recovery clusters, Italy (IT) cohort.'}

suppl_tbl$baseline_clusters_south %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(3, 1, value = as_paragraph('BMI: body mass index, overweight > 25 kg/m\u00B2, obesity > 30 kg/m\u00B2,'), ref_symbols = 'b', part = 'body') %>% 
  footnote(15, 1, value = as_paragraph('Frequent respiratory infections, > 2 per year.'), ref_symbols = 'c', part = 'body') %>% 
  footnote(16, 1, value = as_paragraph('Frequent bacterial infections with antibiotic therapy, > 2 per year.'), ref_symbols = 'd', part = 'body') %>% 
  theme_vanilla

@@

```{r tab-clust-reco-at, tab.cap = 'COVID-19 course and recovery in the survey study participants assigned to the recovery clusters, Austria (AT) cohort.'}

suppl_tbl$cov_clusters_north %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
  footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(4, 1, value = as_paragraph('Self-reported incomplete recovery.'), ref_symbols = 'd', part = 'body') %>% 
  footnote(8, 1, value = as_paragraph('ANX: anxiety.'), ref_symbols = 'e', part = 'body') %>% 
  footnote(9, 1, value = as_paragraph('DPR: depression.'), ref_symbols = 'f', part = 'body') %>% 
  footnote(11, 1, value = as_paragraph('OMH score: overall mental health impairment score'), ref_symbols = 'g', part = 'body') %>% 
  footnote(12, 1, value = as_paragraph('QoL score: quality of life impairment score'), ref_symbols = 'h', part = 'body') %>%
  theme_vanilla

@@

```{r tab-clust-reco-it, tab.cap = 'COVID-19 course and recovery in the survey study participants assigned to the recovery clusters, Italy (AT) cohort.'}

suppl_tbl$cov_clusters_south %>% 
  flextable %>% 
  width(1, width = 3.5, unit = 'cm') %>% 
  width(2:4, width = 3.7, unit = 'cm') %>% 
  width(5, width = 2.6, unit = 'cm') %>% 
  width(6, width = 2.2, unit = 'cm') %>% 
 footnote(1, 5:6, value = as_paragraph('Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. P values corrected form multiple testing with Benjamini-Hochberg method.'), ref_symbols = 'a', part = 'header') %>% 
  footnote(4, 1, value = as_paragraph('Self-reported incomplete recovery.'), ref_symbols = 'd', part = 'body') %>% 
  footnote(8, 1, value = as_paragraph('ANX: anxiety.'), ref_symbols = 'e', part = 'body') %>% 
  footnote(9, 1, value = as_paragraph('DPR: depression.'), ref_symbols = 'f', part = 'body') %>% 
  footnote(11, 1, value = as_paragraph('OMH score: overall mental health impairment score'), ref_symbols = 'g', part = 'body') %>% 
  footnote(12, 1, value = as_paragraph('QoL score: quality of life impairment score'), ref_symbols = 'h', part = 'body') %>%
  theme_vanilla

@@

```{r fig-hact-freq, fig.width = mm_inch(suppl$hact_freq$w), fig.height = mm_inch(suppl$hact_freq$h), fig.cap = 'Frequency of COVID-19 symptoms in the survey study.'}

@@

__Supplementary Figure S\@ref(fig:fig-hact-freq). Frequency of COVID-19 symptoms in the survey study.__

_Frequency of symptoms in first 14 days, $\geq$ 28 days and $\geq$ 3 months after clinical onset of COVID-19 in the Austria (AT, A) and Italy (IT, B) survey study cohorts expressed as percentages of the cohort. Point size and color represents the percentage. Numbers of complete observations are indicated under the plot._

_Dim. appetite: diminished appetite, Imp. concentration: impaired concentration, Imp. walk: impaired walk, Imp. FMS: impaired fine motor skills._

@@

```{r fig-hact-recovery, fig.width = mm_inch(suppl$hact_recovery$w), fig.height = mm_inch(suppl$hact_recovery$h), fig.cap = 'Kinetic of recovery from leading acute COVID-19 symptoms in the survey study.'}

@@

__Supplementary Figure S\@ref(fig:fig-hact-recovery). Kinetic of recovery from leading acute COVID-19 symptoms in the survey study.__

_Percentages of individuals with fever (A), diminished appetite (B), joint pain (C), muscle pain (D), fatigue (E) and tachypnea (F) in the AT (Austria) and IT (Italy) survey study cohorts at particular time points after clinical onset. Numbers of complete observations are indicated under the plots._

@@

_Frequency of symptoms during acute COVID-19 and at the 60-, 100-, 180- and 360-day follow-ups in ambulatory (A), moderate (B) and severe COVID-19 (C) participants expressed as percentages of individuals with the complete longitudinal data set. Point size and color represents the percentage. Numbers of complete observations are indicated under the plots._

@@

_Percentages of individuals with the complete longitudinal data set suffering from smell disorders (A), reduced physical performance (B) and dyspnea (C) in the ambulatory, moderate and severe COVID-19 subsets during acute COVID-19 and at the 60-, 100-, 180- and 360-day follow-ups. Numbers of complete observations are indicated under the plots._

@@

_Association of self-reported and sniffing stick test hyposmia rates was investigated with Cohen's $\kappa$ statistic. Statistical significance was assessed with Wald's Z test corrected for multiple testing with Benjamini-Hochberg method. Percentages of individuals with self-reported and test hyposmia within the entire cohort (A), the ambulatory (B), moderate (C) and severe (D) COVID-19 subsets are presented in bubble plots. Point size and color represents the percentage. $\kappa$ values with 95$\%$ confidence intervals, p values and numbers of complete observations are indicated in the plot captions._

@@

```{r fig-mds-acute, fig.width = mm_inch(suppl$mds_acute$w), fig.height = mm_inch(suppl$mds_acute$h), fig.cap = 'Multi-dimensional scaling analysis of acute COVID-19 symptoms in the survey study.'}

@@

__Supplementary Figure S\@ref(fig:fig-mds-acute). Multi-dimensional scaling analysis of acute COVID-19 symptoms in the survey study.__
_Symptom data for acute COVID-19 (first 14 days after clinical onset) in the Austria (AT) and Italy (IT) survey study cohorts were subjected to two-dimensional multi-dimensional scaling (MDS) with simple matching distance (SMD) between the symptoms. MDS coordinates are presented in point plots. Selected data points are labeled with the symptom names. Percentages of the data set variance associated with the MDS dimensions are indicated in the plot axes. Numbers of complete observations are indicated under the plots._

@@

```{r fig-clust-devel, fig.width = mm_inch(suppl$clust_devel$w), fig.height = mm_inch(suppl$clust_devel$h), fig.cap = 'Definition of the COVID-19 recovery clusters and clustering feature importance in the survey study.'}

@@

__Supplementary Figure S\@ref(fig:fig-clust-devel). Definition of the COVID-19 recovery clusters and clustering feature importance in the survey study.__

_Individuals of the training Austria (AT) study survey cohort were subjected to clustering in respect to symptom-specific recovery times with the PAM (partitioning around medoids) algorithm and Euclidean distance measure._

_(A) Comparison of performance of various algorithms (HCl: hierarchical clustering, SOM + HCl: combined self-organizing map and hierarchical clustering, k-means) and distance statistic in clustering of the training data set investigated by clustering variance (ratio of total between-cluster sum of squares to total sum of squares) and cluster assignment accuracy in 10-fold cross-validation (CV)._

_(B) Determination of the optimal cluster number in the PAM clustering of the training cohort by the bend of the total within-cluster sum of squares curve._

_(C) Permutation importance of the clustering features (symptoms) for clustering of the training cohort expressed as the difference in clustering variance (ratio of total between-cluster sum of squares to total sum of squares) between the initial clustering object and the clustering object with the given variable reshuffled at random._

@@

```{r fig-clustering, fig.width = mm_inch(suppl$clustering$w), fig.height = mm_inch(suppl$clustering$h), fig.cap = 'Clustering of ambulatory COVID-19 individuals by symptom-specific recovery times.'}

suppl$clustering$plot

__Supplementary Figure S\@ref(fig:fig-clustering). Clustering of ambulatory COVID-19 individuals by symptom-specific recovery times.__
